Boosted Regression Trees for Predictive Auto-Tuning

### 2011-12-14 paper/draft1.tex
Minor section name edit
Assorted typo fixes, word changes, XXX removals

Auto-tuning of parameteric, instrumented code has proven to be a highly
effective technique for tailoring an optimal compute kernel for a particular
hardware \cite{XXX}.
% DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
However, a drawback of this approach is that exhaustive, or even thorough,
auto-tuning requires compiling many kernels and calling each one many times,
and this process is slow.  Furthermore, it is desirable to provide users
with unified library abstraction boundaries for operations such as image
filtering and matrix multiplication, even those these operations actually
correspond to a large set of potential problem configurations with a wide
variety of memory access patterns and computational bottlenecks.  How can we
draw on data from previous empirical auto-tuning of related problems on related
hardware to make a just-in-time implementation decision for a novel problem?
This paper presents a machine learning approach to auto-tuning, in which
features of the current hardware platform, the kernel configuration and the
problem instance are passed to a regression model (boosted regression trees)
which predicts how much faster this kernel will be than a reference
baseline.  Combinatorial optimization strategies for auto-tuning that would
normally require evaluating large number of kernel configurations on the real
hardware, are made orders of magnitude faster by evaluating the surrogate
regression model instead.  We validate our approach using the filterbank
correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
0.1 seconds (XXX) of hill climbing on the regression model (which we dub ``{\em
predictive} auto-tuning'') can achieve an average of 95\% of the
speed-up brought by minutes of empirical auto-tuning XXX.  Our approach is not
specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
be applied to almost any templated-code optimization problem, spanning a wide
variety of problem types, kernel types, and platforms.

@@

XXX: use 300\% boost instead of 3x

@@

Even though such a ``black-box'' approach does not try to build an
explicit model per se, it could, if care were taken, help to greatly accelerate
our understanding of the complex interactions between the set of optimizations,
their parameters and the hardware architectural features (XXX: that's what
we are trying to do w/ ML-based, decision trees, etc.).

@@

To alleviate this major weakness, it is intuitively appealing to combine the two
approaches by first constraining the search space with an analytical model and
then exploring the reduced space empirically \citep{chen2005combining,
li2009note}. Unfortunately, this hybrid approach is still bound by the quality
of the analytical model, which remains hard to build by hand.

@@

In this paper, we propose to \emph{learn the model} using non-linear regression
modelling techniques instead of constructing a model manually. By learning the model, one can
hope to achieve elements of the best of both approaches: XXX.
%and capitalize on the large body of
%literature from the machine learning community.

Various statistical prediction techniques have been applied with
success at compile-time for general programs on various CPU architectures
\citep{monsifrot2002machine, stephenson2003meta, yotov2003comparison,
kulkarni2004fast, cooper2005acme, franke2005probabilistic,
hutter2006performance, cavazos2007rapidly, cavazos2008intelligent,
hartono2009annotation, park2011evaluation, fursin2008milepost}.
Relative to this work, our contribution is to show
how to do fast predictive auto-tuning that satisifies the requirements to:
(a) handle the variety of recent multi-core architectures like GPUs \citep{schaa2009exploring},
(b) provide high-performance domain-specific libraries \citep{nukada2009auto, li2009note, kamil2010auto},
(c) that select good implementations at run-time \citep{klockner2011pycuda, pinto+cox:2011gcg}, and
(d) for the full input domain of a library routine \citep{liu2009cross, grauer2011optimizing}.
% XXX: are these last references appropriate
% XXX: there are 4 versions of the stencil work - consolidate to just the full journal version

@@

The heart of predictive auto-tuning is a regression model that acts as surrogate
for a hand-written hardware model or empirical timing estimates.
In our experiments this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
To fit this model, we
form a training set $\mathcal{X}, \mathcal{Y}$ where
each point $x^{(i)} \in \mathcal{X}$ is a tuple $(a^{(i)}, b^{(i)}, c^{(i)})$
and each target $y^{(i)} \in \mathcal{Y}$  reflects the speed of implementation $b^{(i)}$
on inputs $a^{(i)}$ on platform $c^{(i)}$.

@@

@@

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_gflop_scatter_295.pdf}
\includegraphics[scale=.42]{fig_gflop_scatter_480.pdf}
\caption{Computation speed for novel problem configurations using predictive
vs. empirical code specialization.
The scatterplots are roughly symmetric about the axis of equality (main diagonal)
with occasional outliers, indicating that for most problem configurations the
predictive approach gives as good a solution as the empirical approach.
The predictive approach typically requires about 0.1 seconds to suggest an
implementation, whereas the empirical approach requires about 2 minutes.
}
\label{fig:fig_gflop_scatter}
\end{figure}

>>>>

Auto-tuning of parameteric, instrumented code has proven to be a highly
effective technique for tailoring an optimal compute kernel for a particular
hardware.
% DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
However, a drawback of this approach is that exhaustive, or even thorough,
auto-tuning requires compiling many kernels and calling each one many times,
and this process is slow.  Furthermore, it is desirable to provide users
with unified library abstraction boundaries for operations such as image
filtering and matrix multiplication, even those these operations actually
correspond to a large set of potential problem configurations with a wide
variety of memory access patterns and computational bottlenecks.  How can we
draw on data from previous empirical auto-tuning of related problems on related
hardware to make a just-in-time implementation decision for a novel problem?
This paper presents a machine learning approach to auto-tuning, in which
features of the current hardware platform, the kernel configuration and the
problem instance are passed to a regression model (boosted regression trees)
which predicts how much faster this kernel will be than a reference
baseline.  Combinatorial optimization strategies for auto-tuning that would
normally require evaluating large number of kernel configurations on the real
hardware, are made orders of magnitude faster by evaluating the surrogate
regression model instead.  We validate our approach using the filterbank
correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
0.1 seconds (XXX) of hill climbing on the regression model (which we dub ``{\em
predictive} auto-tuning'') can achieve an average of 95\% of the
speed-up brought by minutes of empirical auto-tuning XXX.  Our approach is not
specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
be applied to almost any templated-code optimization problem, spanning a wide
variety of problem types, kernel types, and platforms.

@@

% XXX: use 300\% boost instead of 3x

@@

Even though such a ``black-box'' approach does not explicitly try to build an
interpretable model, per se, it could nonetheless help to greatly accelerate
our understanding of the complex interactions between the set of optimizations,
their parameters and the hardware architectural features.
% (XXX: that's what we are trying to do w/ ML-based, decision trees, etc.).

@@

To ameliorate this weakness, it is intuitively appealing to combine the two
approaches by first constraining the search space with an analytical model and
then exploring the reduced space empirically \citep{chen2005combining,
li2009note}. Unfortunately, such a hybrid approach is still bound by the quality
of the analytical model, which remains hard to build by hand.

@@

In this paper, we propose to \emph{learn} the model using non-linear regression
modeling techniques, rather than constructing a model manually. By learning the model, one can
hope to achieve elements of the best of both approaches. %: XXX.
%and capitalize on the large body of
%literature from the machine learning community.

Various statistical prediction techniques have been applied with
success at compile-time for general programs on various CPU architectures
\citep{monsifrot2002machine, stephenson2003meta, yotov2003comparison,
kulkarni2004fast, cooper2005acme, franke2005probabilistic,
hutter2006performance, cavazos2007rapidly, cavazos2008intelligent,
hartono2009annotation, park2011evaluation, fursin2008milepost}.
Relative to this work, our contribution is to show
how to do fast predictive auto-tuning that satisfies the requirements to:
(a) handle the variety of recent multi-core architectures like GPUs \citep{schaa2009exploring},
(b) provide high-performance domain-specific libraries \citep{nukada2009auto, li2009note, kamil2010auto},
(c) that select good implementations at run-time \citep{klockner2011pycuda, pinto+cox:2011gcg}, and
(d) for the full input domain of a library routine \citep{liu2009cross, grauer2011optimizing}.
% XXX: are these last references appropriate
% XXX: there are 4 versions of the stencil work - consolidate to just the full journal version

@@

The heart of predictive auto-tuning is a regression model that acts as surrogate
for a hand-crafted hardware model or empirical timing estimates.
In our experiments, this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
To fit this model, we
form a training set $\mathcal{X}, \mathcal{Y}$ where
each point $x^{(i)} \in \mathcal{X}$ is a tuple $(a^{(i)}, b^{(i)}, c^{(i)})$
and each target $y^{(i)} \in \mathcal{Y}$  reflects the speed of implementation $b^{(i)}$
on inputs $a^{(i)}$ on platform $c^{(i)}$.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=.9\linewidth]{illustrator/fig_main_R1.pdf}
\caption{
XXX UPDATE THIS WITH (a) (b) (c) XXX
Computation speed for novel problem configurations using predictive
vs. empirical code specialization.
The scatterplots are roughly symmetric about the axis of equality (main diagonal)
with occasional outliers, indicating that for most problem configurations the
predictive approach gives as good a solution as the empirical approach.
The predictive approach typically requires about 0.1 seconds to suggest an
implementation, whereas the empirical approach requires about 2 minutes.
}
\label{fig:fig_gflop_scatter}
\end{figure*}

@@
